---
title: "Diversity Index for Maryland Cities"
author: "Rachel McCrea"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

In this exercise, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

A score of 0 indicates no diversity (everyone is the same race/ethnicity), while a score closer to 100 indicates higher diversity. The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

## Task 1: Load and Describe the Data (3 points)

Run the following code to load the data, then examine it and describe what it contains. Based on the raw numbers alone, which cities stand out as diverse to you?

This dataset has population by race / ethnicity for 128 cities in Maryland. It lists the total population of the city/town, and then breaks the population into 9 racial/ethnic groups. 
Just looking at the raw numbers, it looks like Baltimore is pretty diverse, but that might just be because all of its numbers are bigger than other cities'. Bowie also seems fairly diverse, and Hyattsville, Chillum, and Columbia too. As far as least-diverse towns/cities, Bel Air, Westminster, and Largo all stood out to me. So did Severna Park, Linthicum, and Timonium.
Honestly though, I don't love just looking at the numbers. I think the code will probably tell me something different than what I'm seeing. 

```{r load-data}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/usat_diversity_index.csv")

# Preview structure
View(md_cities)
```

## Task 2: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities |>
  mutate(
    White_pct = White / TotalPop,
    Black_pct = Black / TotalPop,
    Native_pct = Native / TotalPop,
    Asian_pct = Asian / TotalPop,
    NHPI_pct = NHPI / TotalPop
  )
```

## Task 3: Square the Race Percentages

We need to square the race percentages in order to calculate the probability that any two people chosen at random from a given census area are of different races or are Hispanic or not.

```{r square-race}
md_cities <- md_cities |>
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    Native_pct_sq = Native_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```


## Task 4: Calculate Race Probabilities (3 points)

The race probability is the sum the squared probabilities for each of the races. It represents the probability that two people randomly chosen from the same city will be of the same race. Which city has the highest race probability, and why do you think that is?

I ran the code and checked out the dataset, and organized the Race_Prob column from highest to lowest. Looks like Summerfield has the highest race probability at around 87%. I looked through the race percentages, and I think this high race probability is becuase Black residents make up 93% of the population. 

```{r race-prob}
md_cities <- md_cities |>
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + Native_pct_sq + Asian_pct_sq + NHPI_pct_sq
  ) |> 
  arrange(desc(Race_Prob))

View(md_cities)
```


## Task 5: Calculate Ethnicity Percentages and Their Squares (3 points)

Hispanic origin is not a race but an ethnicity or nationality – and a separate Census question. The probability someone is Hispanic or not Hispanic must be calculated separately from race, and Hisp_Prob is the probability that two randomly selected people will be Hispanic or not. That's an important distinction, because unlike race probability, this measures the likelihood of a binary question: that both random people will be Hispanic. What do you notice about the cities with the *lowest* Hisp_Prob scores?

The cities with the lowest Hisp_Prob scores are Chillum, East Riverdale, and Wheaton. They're in PG and Montgomery counties. These cities have some of the largest Hispanic populations in MD by percentage of the city's total population. 

```{r ethnicity}
md_cities <- md_cities |>
  mutate(
    HispTotal = HISPANIC + NotHispanic,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NotHispanic / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  ) |> 
  arrange(Hisp_Prob)

View(md_cities)
```

## Task 6: Multiply Race and Ethnicity Probabilities

Here we multiply the two probabilities together to get the probability that any two random people have the SAME race and ethnicity.

```{r combined-prob}
md_cities <- md_cities |>
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```


## Task 7: Calculate Diversity Index (3 points)

The final step is to calculate the probability that two people randomly chosen WILL NOT BE OF THE SAME RACE OR ETHNICITY. In other words, the highest index scores will be cities where those two random people will be from different backgrounds. Put simply, more diverse. Which cities are the top 3, and where are they located? Look at the bottom of the list - which city is the least diverse, and where is it?

The top 3 most diverse cities in MD are Langley Park, Wheaton, and Adelphi. They all have diversity index scores of 90 or higher (Langley Park has a 97%). Adelphi and Langley Park are both west of UMD, bordering the golf course. Wheaton is in Montgomery County, north of Silver Spring and east of Bethesda. 
The least diverse city in MD is Lake Shore. It's in Anne Arundel County over by Pasadena. 

```{r diversity-index}
md_cities <- md_cities |>
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  ) |> 
  arrange(desc(Diversity_Index))

View(md_cities)
```

### Pitch a story (8 points)

Given the data that you've produced here, write a headline and lede of the most interesting story from this exercise, and add below that details on how you would report it out: who you'd talk to and the questions you'd ask. When you describe the Diversity Index values, do so in a similar way to how USA Today did; the raw values themselves won't mean much to readers.

HEADLINE: Prince George's County has some of the most, least diverse cities in Maryland
LEDE: Maryland's top 10 most diverse cities were all in Prince George's or Montgomery counties, according to the most recent census data. However, Prince George's County was also home to six of the state's ten least diverse cities. 

How I'd report it:
I'd talk to local government in each of these cities, as well as residents. I'd find UMD professors who teach about demographics and diversity, and hopefully get some of the folks who worked on this at USA Today to weigh in. I'd ask people in the different PG cities about how they percieve diversity in their town, about how likely they think it is that they meet someone of a different race/ethnicity of their own that day. Then I'd tell them USA Today's score (explaining it like "there's a 93% chance the next person you meet will be of a different race/ethnicity than you) and see what their take on that is. 

