---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)

```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

I'm seeing a lot of different crime categories, but from a reporting standpoint I think violent crime statistics (assault, homicide, etc) are important to readers, and so are residential break-ins. What really caught my eye  were the auto theft stats: between ~250 and ~450 cars stolen each month with only a few dozen recovered (max 58 recoveries in 1 month). I think there could be a story there. 

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_totals <- pg_crime %>% mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov+ Dec) 
pg_crime_averages <- pg_crime %>% mutate(average_incidents = (Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec)/12) 


```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The categories with the highest average number of monthly incidents are accidents (and accidents with impounds), stolen vehicles, car thefts (stealing things from cars?), and thefts. Lots of car-related crimes, and PG County seems to have a car stealing issue if there's an average of 399 incidents per month. That being said, we'd probably have to compare the number of stolen cars to the total population in PG county.
Homicide is the category with the lowest monthly average, but we'd get more context on that number (5.8 per month) if we compared that to total county population and to the monthly homicide average from other Maryland counties. Other categories with low averages include breaking and entering that isn't commercial or residential, assault shootings, assault weapons, and sex offenses. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages %>% mutate(june_change = (Jun - average_incidents)/average_incidents * 100) %>% mutate (july_change = (Jul - average_incidents)/average_incidents *100 ) %>% mutate (august_change = (Aug - average_incidents)/average_incidents * 100)

```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

I'm hoping I did this right -- auto thefts went up in the summer, but so did recoveries in june and july. Commercial robberies dropped, then shot up significantly in August. There were more sex offenses and car robberies in June and July, then both dropped in August. Thefts went down. For a lot of crimes, it looks like the number of incidents went up in June and July and dropped in August. So the theory is kind of right and kind of wrong, and it depends on the crime. 


### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Almost 400 cars per month were stolen on average in Prince George's County last year.

It was the second most common incident  listed in the county's crime data, after car accidents. Five vehicles were stolen per every one thousand people in 2024, and only about a tenth of stolen cars were recovered by police. Car break-ins were almost as common with 391 per month. 

```{r}
# 947,430 is the county population according to the Census Bureau 
947430 / 4790
4790 / 947430
# google says to find rate per 1000 i use 4790 / 947430 * 1000
(4790 / 947430) * 1000
# MAN i hope that's right. 
(4703 / 947430) * 1000 
# ^ for car break-ins
```


### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
