---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mmstats <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```
```{r}
glimpse(mmstats)
```




### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.

```{r}
sample5 <- mmstats %>% 
  sample_n(5)
sample10 <- mmstats %>% 
  sample_n(10)
```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.
```{r}
sample5 %>% summarize(mean = mean(red), sd = sd(red))
sample10 %>% summarize(mean = mean(red), sd = sd(red))
#sample5 red: mean = 70.8, sd = 14.23728 
#sample10 red: mean = 78.5, sd = 20.09007
#yikes
```


### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4
```{r}
mmstats %>% 
  summarize(mean = mean(red), sd = sd(red))
#population red: mean = 72.98, sd = 16.45808
```


### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

The mean and standard deviation of red in the population was 72.98 and 16.46. The sample of ten wasn't accurate because it had a standard deviation of 20 and a mean of 78. The sample of 5 also wasn't accurate because it predicted a mean of 70 (closer than 78) and a standard deviation of about 14 (closer than 20!). Weird that the smaller sample got closer to the actual population than the larger sample.  

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The samples I took weren't super accurate compared to the actual mean and standard deviation of the population. I think that means I wasn't taking large enough samples, or that sometimes random sampling can get you weird data because there's always the chance of getting a sample that isn't super representative. I think it also means there's a pretty high amount of variability in the m&ms dataset, the standard deviation seems fairly high. 



When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.