---
title: "HW8_GDP_DJIA"
---

What is the relationship between the stock market, represented by the Dow Jones Industrial Average, or DJIA, and the Gross Domestic Product, a measure of the U.S. economic activity?

You will examine that question for the past decade, from 2013 through 2024.

The U.S. economy in 2013 had begun to emerge from the Great Depression and heated up in the years before the COVID-19 pandemic hit. The government and private sector shutdown in a bid to save lives, ease overcrowded hospitals and prevent a wider spread of a little-known but often debilitating and deadly virus. As economic life returned to pre-pandemic levels, the outlook for the economy mostly was positive, but inflation remained high for several years.

Your task in this assignment will be to examine the relationship between those two key economic indicators, the DJIA and GDP, for the past decade. You will calculate a correlation coefficient and create a scatterplot to visualize the relationship between the two variables. As usual, you will use the Tidyverse to do this work.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the GDP and DJIA data

Load the GDP data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/gdp_djia.csv

```{r}
gdp_djia <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/gdp_djia.csv")
```

### Task 3: Calculate the correlation coefficient

Use the `cor()` function to calculate the correlation coefficient between the DJIA and GDP.

```{r}
gdp_djia %>% summarize(correlation = cor(GDP, DJIA, method = "pearson"))
#0.953...
#pretty high correlation coefficient
```

### Task 4: Create a scatterplot

Use the `ggplot()` function to create a scatterplot of the DJIA and GDP. Be sure to label your axes and add a title.

```{r}
gdp_djia %>% 
  ggplot() +
  geom_point(aes(x=GDP, y=DJIA))+
  geom_smooth(aes(x=GDP, y=DJIA), method="lm")+
  geom_text(aes(x=GDP, y=DJIA, label=Quarter))

#I'M ASKING CHATGPT HOW TO ADD A TITLE!

#Me: hey man, I'm working with some GDP and DJIA data in R and want to add a title to my scatterplot. How do I do that? 
#ChatGPT: Hey! You can easily add a title to your scatterplot in R using the title() function or by using the ggtitle() function if you're using ggplot2. Here's how you can do it in both cases: [chatgpt gave me some code] 
#Me: Great, I'm using ggplot and will try the title() function. 
#ChatGPT: Actually, just to clarify, when using ggplot2, the title() function isn't the right one to add a title. You should use ggtitle() instead, like in the second example I gave. Here's how you would do it in ggplot2:
        
        #R
        #Copy
        # Load ggplot2
        #library(ggplot2)
        
        # Example data
        #data <- data.frame(
         # gdp = c(2.3, 2.4, 2.5, 2.6, 2.7),
          #djia = c(11000, 11200, 11500, 11800, 12000)
        #)
        
        # Create scatterplot with title
        #ggplot(data, aes(x = gdp, y = djia)) +
         # geom_point() +
          #ggtitle("Scatterplot of GDP vs DJIA") +  # Add title here
          #xlab("GDP Growth") +
         # ylab("DJIA")
        #If you try title() in ggplot2, it won’t work. Just use ggtitle() for the title, and you'll be good to go!
        #Let me know if you run into any other issues!

#I don't think we're in ggplot2 but I'm gonna give ggtitle() the old college try anyway. 
gdp_djia %>% 
  ggplot() +
  geom_point(aes(x=GDP, y=DJIA))+
  geom_smooth(aes(x=GDP, y=DJIA), method="lm")+
  geom_text(aes(x=GDP, y=DJIA, label=Quarter))+
  ggtitle("GDP vs DJIA, 2013-2024")
#YOOOOO THAT WORKED YAY!!!
```

### Task 5: Describe our results

Write a few sentences about the relationship between the DJIA and GDP. Is there a strong correlation between the two variables? What does the scatterplot show?

I feel confident saying there's a strong positive relationship between the DJIA and the GDP. The scatterplot indicates this -- you can tell from the shape of the data points as well as the line of best fit. The correlation coefficient for this data is 0.9531058, which is close to 1 and indicates a strong correlation. I'd say the GDP and DJIA are pretty closely tied, though there might be a third variable influencing this relationship. 


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

