---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "YOUR NAME"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

It looks like the higher the FARMS percentage, a lower percentage of students met or exceeded math expectations. This is a lot different from MoCo PA, where there was pretty much no correlation at all between FARMS and math performance. 

### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```
P-value is SUPER low. Adjusted R^2 is 0.67, which I think means FARMS percentage impacts 67% of the data. The min is -28 and the max is 29, so we've got some pretty big residuals I think. The median is pretty close to zero and the quarters are around 6. 
I'd say this model would be good, but not fully accurate, at predicting math scores based on FARMS percentage. 

Provide the model's output for the three statistics mentioned in the class activity.

P-value = less than 0.00000000000000022
Adjusted R-squared = 0.6734
Min residual = -28.0844


Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

I think this is worth examining further. It looks like FARMS percentage could be a big predictor of math performance, and there needs to be more exploration into whether this is correlation or causation, and why the two variables move together. 

### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

The biggest differences between predicted result and actual math scores are 55 (Stedwick Elementary) and 56 (Wayside Elementary). 
Stedwick Elementary's zip code has a high Hispanic population, a median income that's a little less than the average in Maryland, and a poverty rate that's 20% higher than Maryland's average. It looks like there are a lot of families living in this area, and a lot of single-family homes. The most interesting stat to me is that it has over double the percentage of foreign-born residents than the rest of Maryland. It looks like there are a lot of immigrants in this neighborhood from Latin America.

Wayside Elementary is in Potomac. It's a majority-white neighborhood with a lot of teenagers, and a household income of 250K. That's over twice the state median income. There are also a lot of families in this area, which is mostly single-unit housing. Like Stedwick's zip code, there are a lot of foreign-born residents (almost double the rest of Maryland). The majority of immigrants to this neighborhood are from Asia. I think the most interesting stat is that the percentage of people with a bachelor's degree is  double the percentage for the rest of Maryland. 



## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

I think I'd draw the conclusion that students at schools in richer areas perform better at math. If the school board member said, like in PA, that schools with fewer students on FARMS do better on standardized math tests, I'd probably say the data backs that up. I think that any story pitch would have to be for a deep dive into this -- use the school board member's comment as a jumping-off point and dig into the relationship between poverty and math scores in MoCo. I'd do a lot more reporting to look at differences between schools with more/less students on FARMS, higher/lower math scores, and I'd check out the outliers and what makes them different. Would need to go into the "why" of this! 