---
title: "FINAL_FINAL_PROJECT_MCCREA"
author: "Rachel McCrea"
date: "2025-05-16"
output: html_document
---
ANALYZING HOSPITALS PER COUNTY, POPULATION PER COUNTY, AND PER CAPITA PERSONAL INCOME PER COUNTY. MY INITIAL HYPOTHESIS WAS THAT RICHER, MORE POPULOUS COUNTIES WOULD HAVE MORE HOSPITALS. 


---------------------------------------- JOINING DATA -------------------------------------------------- 

Link to the AI conversations I had: https://github.com/rachmccrea/jour405_spring25/blob/main/project/Troubleshooting%20right%20join%20error%20in%20R%20with%20GEOID.txt

Loading in the data again (this is copied from the bottom of the notebook, which has notes on what I got help from AI on, and all the stuff we needed for one of the earlier project assignments): 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(tidycensus)
library(knitr)
library(ggplot2)
library(ggrepel)
library(sf)

```
```{r}
hospitals <- read_csv("md_hospitals_state_data.csv")
hospital_by_county <- read_csv("hospital_by_county.csv")
census_api_key("738bbf398565b3bd6d468669a0ef4374f13c150b", overwrite=TRUE)
population <- get_decennial(geography = "county", 
                            variable = "P1_001N", 
                            state = "MD",
                            year = 2020, 
                            sumfile = "pl")
income <- read_csv("md_pcpi_bea.csv")
```
Added  GEOID to hospital_by_county using population, went in thru excel and making those changes. 
Ended up putting in the GEOID numbers by hand, which introduces the risk of human error. Added Queen Anne's and Caroline Counties, which don't have hpsitals. 

Loading updated hospital_by_county data

```{r}
write_csv(hospital_by_county, "hospital_by_county_to_edit.csv")
hospitals_by_county <- read_csv("hospital_by_county_geoid.csv")
```
Need to convert one of them from chr to dbl 

```{r}
glimpse(population)
glimpse(hospitals_by_county)
#ah it's population. 

population <- population %>% 
  mutate(GEOID = as.numeric(GEOID))

glimpse(population)
glimpse(hospital_by_county)
#BAM i did that from MEMORY let's GO
```

Now I try to join them
```{r}
joined_hospital_population <- hospitals_by_county %>% 
  right_join(population, join_by("GEOID"=="GEOID"))
#YAY IT WORKED 
```

```{r}
joined_hospital_population <- hospitals_by_county %>% 
  right_join(population, join_by(GEOID==GEOID))

joined_hospital_population %>% 
  clean_names()

view(joined_hospital_population)
#WAHOO!! 

```
Ok, so now I know that 2 counties don't have hospitals, and I have the number of hospitals per county and county population in the same dataset.

Now I want to join number of hospitals with income. After that, I'll join all three. That'll leave me with 3 good datasets to work with: hospitals+population, hospitals+income, and hospitals+population+income. 


```{r}
#step 1: write income csv to folder, remove maryland row, add GEOID, and read it back in. 
#deleted write_csv() because when it ran it would delete the GEOIDs I added. 
income_per_capita <- read_csv("income.csv")

#step 2: join to hospital_by_county 

joined_hospital_income <- hospitals_by_county %>% 
  right_join(income_per_capita, join_by("GEOID"=="GEOID"))
#consulted terpAI because this wasn't running, and then realized that when I did the run-all before turning this in I destroyed the version of income.csv that had the GEOIDs. went in, fixed it, got rid of the write_csv command. 

read_csv("joined_hospital_income.csv")
#had to go into excel and fix the dataset - mixed up baltimore city and county but it should be good to go now. 

#ok there we go. 
```


Now we join all three! 

```{r}
joined_all <- joined_hospital_population %>% 
  right_join(income_per_capita, join_by (GEOID==GEOID))

#cutting out the columns we don't need right now 

#joined_all_simplified <- joined_all %>% 
  #select(GEOID, number_of_hospitals, population, COUNTY, 2023_PCPI, RANK_PCPI_2023)
#^^this didn't work so I'm commenting it out. 

#aaaand i need to change some column names. going back to excel. 


joined_all <- read_csv("joined_all.csv")
#changed some column names in excel, deleted write_csv because it wrecked my changes on my last run-through. 

#let's try again. 

joined_all_simplified <- joined_all %>% 
  select(GEOID, number_of_hospitals, county_population, COUNTY, PCPI_2023, RANK_PCPI_2023)
#YAY
```



------------------------------ DESCRIPTIVE / SUMMARY STATISTICS ----------------------------------------

Step 1: I want to know the summary stats for number of hospitals in a county so I can see variation across Maryland.
```{r}

#it was at this point I realized that the # of hospitals for QA and Caroline were NA. Time to change that. 

#write_csv to excel to change it 

write_csv(joined_all, "joined_all_to_fix.csv")
#making a copy, adding county name and 0 instead of NA to the # of hospitals section, reading back in 
#learning how to use write_csv the right way! I'm not sure how to fix my write_csv issues from the last step but I started making copies so I don't blow things up on the final Run All!
joined_all_complete <- read_csv("joined_all_complete.csv")

#summary stats!
joined_all_complete %>% 
  summarize(mean = mean(number_of_hospitals), min = min(number_of_hospitals), max = max(number_of_hospitals), sd = sd(number_of_hospitals))

#how many hospitals are in MD? 
joined_all_complete %>% 
  summarize(sum = sum(number_of_hospitals))
```
There are 64 total hospitals in Maryland, according to this data set. The mean number of hospitals in a county is 2.67, which we could round up to 3. The minimum number of hospitals is zero, and the max is 15. The standard deviation is 3.27. 
From these summary statistics, I can tell there's a lot of variation in the number of hospitals in a county/jurisdiction. 

Step 1.5: I think it's worth doing a data visualization to see how hospitals are distributed across Maryland. 


```{r}
ggplot()+
  geom_bar(data = joined_all_complete, aes(x = county.x, weight = number_of_hospitals))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #learned how to do that 45 degree thing for my 472 project using terpAI or chat, but I don't think I exported the conversation :( it gave me the theme(axis.text.x) stuff. 
  labs(
    title="Number of Hospitals Per County in Maryland", 
    x="County",
    y="Number of Hospitals"
  )

```
This distribution is interesting! Four counties/jurisdictions have way more hospitals than the rest, and they're all in and around Baltimore or DC. Should do the same visualization for population and see if these counties have huge populations. Looks like a lot of counties only have one or two hospitals, and maybe the mean was skewed in my summary statistics by these four hospital-heavy counties. I think it's interesting that Dorchester, Washington, and Wicomico have 3 hospitals each even though they're not big population centers the way the top four counties are. 

Step 2: I want to know summary stats for population so I can see variation. 

```{r}
joined_all_complete %>% 
  summarize(mean = mean(county_population), min = min(county_population), max = max(county_population), sd = sd(county_population))

```
The mean county population is about 257K and the standard deviation is 315K. The minimum populatino is only 19K (I checked, that's Kent) but the max is over a million (Montgomery). Like the number of hospitals in a jurisdiction, there's a lot of variation in county population in Maryland. 

Step 2.5: Data viz for population -- does it match # of hospitals? 

```{r}
ggplot()+
  geom_bar(data = joined_all_complete, aes(x = county.x, weight = county_population))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #learned how to do that 45 degree thing for my 472 project using terpai or chat
  labs(
    title="County Population in Maryland", 
    x="County",
    y="Population"
  )

```
This is interesting! Montgomery and PG are the two largest population centers in Maryland by a significant amount. Baltimore County also has a large population. However, Baltimore City is almost tied in population with Anne Arundel, but Anne Arundel only has two hospitals and Baltimore has 15. Why are there so many hospitals in Baltimore but comparatively few in Anne Arundel? 


Step 3: Summary stats for income so I can see variation -- that's going to be interesting. 

```{r}

joined_all_complete %>% 
  summarize(mean = mean(PCPI_2023), min = min(PCPI_2023), max = max(PCPI_2023), sd = sd(PCPI_2023))

```
The mean per capita personal income in 2023 was 69K. The county with the minimum PCPI was Somerset with 37K, and the max was over 100K in Montgomery. 

Step 3.5: Bar chart for county PCPI -- are the counties with the highest number of hospitals the wealthiest? 

```{r}

ggplot()+
  geom_bar(data = joined_all_complete, aes(x = county.x, weight = PCPI_2023))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #learned how to do that 45 degree thing for my 472 project using  terpai or chat
  labs(
    title="Per Capita Personal Income by County in Maryland ", 
    x="County",
    y="2023 per capita personal income"
  )

```

This doesn't have the huge differences we saw in population and hospital distribution. Montgomery does have the highest PCPI in the state, but PG county has a much lower one and is still one of the counties with the highest population and most hospitals. Neither Baltimore city or county stand out as particularly rich, either, though they're still on the high end. Meanwhile, Talbot and Harford are almost as wealthy as Montgomery but only have one or two hospitals each and pretty small populations. 


----------------------------------------- RUNNING TESTS ------------------------------------------------
Step 1: Linear regression models 

Starting with a linear regression with population and hospital numbers. I feel like that's a good thing to do right off the bat to get some scatterplots and p-values and residuals. 
I'm gonna do basically the same thing we did in class: copy in the ggplot code to make a scatterplot, then copy in the model code to run the regression. 

```{r}
joined_hospital_population |>
  ggplot(aes(x = value, y = count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "county population", y = "number of hospitals", title = "number of hospitals per county by population")
```
Looks like some counties with high populations have a high number of hospitals, but Baltimore City is an outlier with 15 hospitals. 
There's also a county with a relatively high population ond only a couple hospitals. Some with much lower populations have 3 instead of 2. 


```{r}
model <- lm(count ~ value, data = joined_hospital_population)

summary(model)
```
Ok looking through this: 
P-value looks good (had to google what a statistically significant p-value was, forgot it was 0.05)
Adjusted R-squared is 0.42...does that mean it accounts for 42% or 0.42%? Had to google this too, looks like the closer it is to 1 the better. Sooooo it's looking like population Kinda But Not Totally affects the number of hospitals in a county so far. 
Min = -3
Max = 9 
Not liking those residuals. 
My interpretation: population explains some, but not all, of the number of hospitals in a county. 


Checking to see whether I get a different result if I switch the column order in the code: 

```{r}
model <- lm(value ~ count, data = joined_hospital_population)

summary(model)
```
I think I probably made the right call with the other model. this one is a little confusing -- the big population numbers are hard and the p-value and adjusted r squared are the same. 


Now I'm gonna run the same thing for hospitals and income. 

```{r}
#changed column names in excel
joined_hospital_income <- read_csv("joined_hospital_income.csv")

joined_hospital_income |>
  ggplot(aes(x = PCPI_2023, y = number_of_hospitals)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "per capita personal income", y = "number of hospitals", title = "number of hospitals per county by per capital personal income ")
```


That doesn't look like much of anything! Maybe a slight positive slope, but that's it. 

Running the linear regression model!

```{r}
model <- lm(number_of_hospitals ~ PCPI_2023, data = joined_hospital_income)

summary(model)

```
P-value is 0.8 and adjusted r^2 is -0.0437. Max residual is huge. Looks like I can disregard this. 

But I'm gonna try the same thing using rank_pcpi instead of pcpi_2023 to see how the richest and poorest counties compare. See if there's anything there. 


```{r}
joined_hospital_income |>
  ggplot(aes(x = RANK_PCPI_2023, y = number_of_hospitals)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "per capita personal income", y = "number of hospitals", title = "number of hospitals per county by per capital personal income ")
```


```{r}
model <- lm(number_of_hospitals ~ RANK_PCPI_2023, data = joined_hospital_income)

summary(model)

```

Once again, neither of these tests seems to point to a relationship between county wealth and number of hospitals. 

--------------------------------------------------------------------------------------------------------

Step 2: Multiple regression!

I want to see how population and income interact to impact the number of hospitals in a county. The linear regressions didn't show much of an individual impact from either of these variables, but a multiple regression can add a new angle/some more context. 


```{r}
multiple_regression <- lm(number_of_hospitals ~ county_population + PCPI_2023, data = joined_all_complete)
summary(multiple_regression)
```
Alright looking at the numbers here, it's not looking great. The adjusted r-squared is 0.43, which means that income and population have a fairly small impact on the number of hospitals in an area. That means there are other variables in play here! The residuals are also kind of big, and the coefficients are small. This makes it seem less likely that income and population are the main determinants of the number of hospitals in a county. 


--------------------------------------------------------------------------------------------------------


Step 3: looking at distribution of hospitals again. 

I'm interested in the distribution of hospitals, since it looks like the 2 counties without any are both on the Eastern Shore. How are hospitals distributed between the Shore and Western MD? 

```{r}
#I based this code off a lab from 472
joined_hospital_population <- joined_hospital_population %>% 
  mutate(shore = case_when(
    NAME  == "Caroline County, Maryland" ~ "shore",
    NAME == "Cecil County, Maryland" ~ "shore", 
    NAME == "Dorchester County, Maryland" ~ "shore", 
    NAME == "Kent County, Maryland" ~ "shore", 
    NAME == "Queen Anne's County, Maryland" ~ "shore", 
    NAME == "Somerset County, Maryland" ~ "shore", 
    NAME == "Talbot County, Maryland" ~ "shore", 
    NAME == "Wicomico County, Maryland" ~ "shore", 
    NAME == "Worcester County, Maryland" ~ "shore", 
    TRUE ~ "not_shore"
  ))
```

```{r}
#joined_hospital_population %>% 
  #group_by(shore) %>% 
  #summarize(hospitals = sum(number_of_hospitals))

#"shore" is coming up NA so I'm going to put "0" in for "NA" in excel for the two counties missing hospitals. is that allowed? also putting the county names in where they say "NA." 

joined_hospital_population <- read_csv("joined_hospital_population.csv")

#trying again 
joined_hospital_population%>% 
  group_by(shore) %>% 
  summarize(hospitals = sum(number_of_hospitals))
#shore has 11 hospitals, not_shore has 53. 
#population now 
shore_simple <- joined_hospital_population%>% 
  group_by(shore) %>% 
  summarize(hospitals = sum(number_of_hospitals), population = (sum(county_population)))

shore <- joined_hospital_population %>% 
  mutate(number_of_hospitals = as.numeric(number_of_hospitals)) %>% 
  mutate(total_population = sum(county_population), total_hospitals = sum(number_of_hospitals)) %>% 
  mutate(pct_population = county_population/total_population, pct_hospitals = number_of_hospitals/total_hospitals) %>% 
  mutate(hospitals_per_1000 = number_of_hospitals/county_population * 1000)
#getting an error message. asking terp ai. 
#looks like i forgot to change "population" to "county_population" in the last line. thanks terp ai!

glimpse(shore)

ggplot()+
  geom_bar(data = shore_simple, aes(x = shore, weight = hospitals))+
  labs(
    title="Hospitals on the Eastern Shore vs. Mainland MD", 
    x="Region",
    y="Number of Hospitals"
  )

ggplot()+
  geom_bar(data = shore_simple, aes(x = shore, weight = population))+
  labs(
    title="Population on the Eastern Shore vs. Mainland MD", 
    x="Region",
    y="Population"
  )

```


If I did this right, that means the shore has 7% of MD's population but 17% of its hospitals. Non-shore counties have ~92% of the population and 82% of hospitals. Looks like the shore has comparatively more hospitals than the rest of maryland, even though a couple counties don't have one. When I look at the data visualizations, it doesn't look like a huge difference between population proportion and hospital proportion. 


----------------------------------------------------------------------------------------------------

Step 4: Correlation 

Derek suggested running a correlation test, which turned out to be a great idea. The correlation between population and hospitals is much, much higher than the correlation between income and hospitals. This backs up a lot of what my descriptive stats were telling me: that a lot of hospitals were located around population centers. 

```{r}
joined_all_complete %>% summarize(correlation = cor(county_population, number_of_hospitals))
#correlation = 0.688623 -- higher than expected!

joined_all_complete %>% summarize(correlation = cor(PCPI_2023, number_of_hospitals))
#correlation = 0.1204596 -- lower than expected!

```
--------------------------------------------------------------------------------------------------------

Step 5: Z-Scores

How big of an outlier is Baltimore with its 15 hospitals? Let's find out! 

```{r}
# z-scores
joined_z_scores <- joined_all_complete |>
  mutate(
    mean_hospitals = mean(number_of_hospitals),
    sd_hospitals = sd(number_of_hospitals),
    z_score = (number_of_hospitals - mean_hospitals) / sd_hospitals
  )
view(joined_z_scores)

# unusual counties
unusual_counties <- joined_z_scores |>
  filter(abs(z_score) > 2) |>
  arrange(desc(abs(z_score)))

# display
cat("unusual hospital counties:", nrow(unusual_counties), 
    "out of", nrow(joined_z_scores), "total counties (", 
    round(nrow(unusual_counties)/nrow(joined_z_scores)*100, 2), "%)")

```
According to this Z-score data, Baltimore City is coming up on FOUR standard deviations above the mean, so it's a huge outlier. It also seems to be the only one with an "unusual" number of hospitals.  

------------------------------------STORY PITCH---------------------------------------------------------


PTICH: What makes Baltimore a hospital hub? 
This would be a deep dive into the different hospitals in Baltimore, what they specialize in, and how the hospital industry impacts the city. 

MAIN FINDINGS: My original hypothesis was that hospitals would be concentrated in counties that have a high population and high income levels (in essence, that access to healthcare correlated with wealth and population). That turned out to not be true. Though there was some correlation between population and hospitals, and there were more hospitals in and around the DC and Baltimore population centers, there wasn't much correlation between hospitals and income. My linear and multiple regressions, and my correlation coefficient, led me to that finding. If I had to choose another story to write, I'd probably write one debunking my previously held idea that wealthier areas had better healthcare acces -- it's not necessarily true according to my analysis. 
However, I did find that Baltimore has way, way more hospitals than any other jurisdiction in Maryland. The 15 hospitals in the city made it an extreme outlier, according to the z-score test that I ran. Conventional knowledge tells me that it makes sense that there would be more hospitals in a big city, but Baltimore isn't THAT big population-wise and its z-score is pretty significant. 
That leads me to my main pitch: what does the hospital/healthcare industry in Baltimore look like? Why are there so many more hospitals in the city than anywhere else? Is it a hub for a certain kind of treatment, or are there lots of medical schools in the city? What's the history of healthcare in Baltimore? Does it have a reputation in medical circles? How are hospitals distributed throughout neighborhoods -- does every neighborhood have easy access to a hospital? There's so much to dive into here. 
POTENTIAL PROBLEMS: My findings were pretty broad. Potential problems could come from a lack of specificity with my analysis, or I could be seeing things that aren't there: maybe all big cities have a disproportionate number of hospitals. Since this is such a broad topic, I could probably continue analyzing data in different ways forever and still not know if I've found the best story to write. 

REPORTING PLAN: What are my next steps? who do I talk to? where do I go? 
My immediate next step would be to do some background research. I'd research the different hospitals, their locations, their history, their specializations, who runs them (university? religious org? state?) and any other important information. Next up would be reaching out to one or two academics at the UMD School of Medicine in Baltimore and the medical school at Johns Hopkins. The hope would be that a working professional and academic working at one of these schools could provide some more context on what the medical scene in Baltimore looks like, and could be a jumping-off point to talk to other sources. 
Throughout the reporting process, I'd also like to talk to city officials and the Baltimore chamber of commerce to get their view on the impact of all these local hospitals on the city. I'd also want to try to figure out what these hospitals specialize in, talk to the doctors involved, and maybe even shadow one or two. I'd like to go to neighborhoods with many hospitals or very few hospitals to talk to residents about the impact of hospitals (or the lack of hospitals) on their neighborhood. 