---
title: "final_project_mccrea"
author: "Rachel McCrea"
date: "2025-04-19"
output: html_document
---

TESTS I COULD RUN: 

From my own brain: 
- Calculate the hospitals per 1000 rate
- Data visualization of hospitals per capitia, plus their locations (map). Not a test but still useful.
- Chi-square test - I think the data I have is applicable
- ANOVA test - I think the data I have is applicable 
- Would use both chi-square and ANOVA to test my hypothesis: that wealther counties have a higher hospital rate.

When I asked terp ai (conversation copy-pasted below), it told me to run the correlation coefficient (should've thought of that) and a linear regression. Both sounds good to me, so I'll probably use them too. 

Terp AI conversation: 

You  

Hey dude. I hope your day is going well. I'm working with R and the tidyverse to analyze some data: the number of hospitals per capita in Maryland counties and the median income of those counties. Any idea what kinds of tests I could run for this? I've already thought about chi-square and anova tests.
Logo
GPT 4o-mini - UMD3  

Hello! I'm glad to assist you with your data analysis in R and the tidyverse. Your project sounds interesting! Here are some statistical tests and methods you could consider for analyzing the relationship between the number of hospitals per capita and the median income of Maryland counties:

1. Correlation Analysis
Pearson Correlation Coefficient: This test measures the strength and direction of the linear relationship between two continuous variables (e.g., hospitals per capita and median income).
Spearman's Rank Correlation: If your data is not normally distributed, this non-parametric test can assess the relationship between the two variables.
2. Regression Analysis
Linear Regression: You can model the relationship between the number of hospitals per capita (dependent variable) and median income (independent variable). This will help you understand how changes in income relate to the number of hospitals.
Multiple Regression: If you have additional variables (e.g., population density, urban vs. rural classification), you could include them to see how they affect the relationship.
3. ANOVA (Analysis of Variance)
One-Way ANOVA: If you categorize the counties based on income brackets (e.g., low, medium, high), you can use ANOVA to test if there are significant differences in the number of hospitals per capita across these groups.
4. Chi-Square Test
This test is more suitable for categorical data. If you can categorize your counties based on the number of hospitals (e.g., low, medium, high) and median income, you can explore the association between these categories.
5. Non-parametric Tests
If your data does not meet the assumptions required for parametric tests (e.g., normality), you might want to consider non-parametric alternatives like the Kruskal-Wallis test for comparing more than two groups.
6. Visualization
Scatter Plots: Visualize the relationship between hospitals per capita and median income.
Box Plots: If using ANOVA, box plots can help visualize the distribution of hospitals per capita across different income categories.
Example Code in R
Here’s a simple example of how you might perform a linear regression in R:

R
# Load necessary libraries
library(tidyverse)

# Assuming your data frame is called 'data'
# with columns 'hospitals_per_capita' and 'median_income'

# Linear regression
model <- lm(hospitals_per_capita ~ median_income, data = data)
summary(model)

# Visualize the relationship
ggplot(data, aes(x = median_income, y = hospitals_per_capita)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Hospitals per Capita vs. Median Income",
       x = "Median Income",
       y = "Hospitals per Capita")
Conclusion
These tests will help you understand the relationship between hospitals per capita and median income. Make sure to check the assumptions of each test you choose to use. If you need further assistance or specific examples, feel free to ask!

You  

Thanks!
Logo
GPT 4o-mini - UMD3  

You're very welcome! If you have any more questions or need further assistance with your analysis, feel free to ask. Good luck with your project, and have a great day! 😊




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(tidycensus)
library(knitr)
```

Hospital data 
```{r}
hospitals <- read_csv("md_hospitals_state_data.csv")

hospitals %>% 
  group_by(County) %>% 
  summarize(count = n())
```
This CSV has a list of hospitals in MD with their name, county, address, and type. My goal with this data is to have each county be a column including the number of hospitals, county population, and per capita rate of hospitals. Would love some guidance on where to go from here with this this dataset -- I started by using group_by and summarize to figure out how many hospitals per county. Next, I think I'd try saving that as a new dataset and then joining it with population data. I'm gonna try that at the bottom of this notebook once I load population data. 

Population data

```{r}
census_api_key("github doesn't get my api key", overwrite=TRUE)
```
```{r}
population <- get_decennial(geography = "county", 
                            variable = "P1_001N", 
                            state = "MD",
                            year = 2020, 
                            sumfile = "pl")
#did a tidycensus lab a couple weeks ago in Ryan Little's data class. 
#used walker-data.com to figure out how to pull decennial census
#used chatgpt when i was having trouble finding the variable to pull total population per county. told it what i was trying to do, and that i couldn't find the variable. it gave me the variable and told me how to structure my code, specifiying the state and putting sumfile = pl. 
```
This dataset has census data listing the total population of each MD county. I plan to use it to calculate the hospital rate per capita, next step is to modify the hospital csv and join it with this and start doing some serious coding. 

Per capita personal income data 
```{r}
income <- read_csv("md_pcpi_bea.csv")
#did some editing in excel - changed the row names, got rid of the header, deleted data from other states, turned it into a csv. '
```
This is per capita personal income data by county from 2021, 2022, and 2023. The 2023 data is probably going to be the most relevant to my project. I expect to spend most of the time working with the 2023_PCPI column. I wish it also had a median income measure, I think that would protect against outliers messing with the data. 


Life expectancy data 
```{r}
life_expectancy <- read_csv("life_expectancy_md.csv")
```


This csv is just life expectancy by county, plus a row that has life expectancy for the whole state. This came from a much larger excel spreadsheet also saved to my project folder. 

TRYING TO JOIN HOSPITAL AND POPULATION DATA (need help)
```{r}
hospitals <- read_csv("md_hospitals_state_data.csv")

hospital_by_county <- hospitals %>% 
  group_by(County) %>% 
  summarize(count = n())

write_csv(hospital_by_county, file = "hospital_by_county.csv")
#learned how to do that ^ in data journalism this week 

#cleaned up county and row names in excel

hospital_by_county <- read_csv("hospital_by_county.csv")

joined_hospital_population <- hospital_by_county %>% 
  right_join(population, join_by(county==NAME))
#learned how to join in data class. None of the joins I tried preserved both the number of hospitals and the population -- it was either one or the other and I'm stuck. 
```


